Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-09-08 prop_letter.tex
prop_letter.tex Updating Andy's note...

We also investigate, though with less insight-generating results, the
second key aspect of imprecise quantum measurement, which is its
statistical nature as reflected in the application of the Born rule
determining the probability of a given measurement
outcome~\citep{Born1983,peres1995quantum,544199}.  Here, Gleason's
theorem establishes that there is no alternative to the Born rule by
demonstrating, using reasonable continuity arguments in Hilbert spaces
of dimension $d\ge3$, the existence of a unique state $\rho$
consistent with the statistical predictions computed from that
rule~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.  Our
question then concerns what happens when infinite precision cannot be
achieved.  We are able to show that, while a QIVPM incorporating the
effects of finite precision might not be consistent with Gleason's
unique state~$\rho$ on all projectors defined on a Hilbert space
$\Hilb$ of dimension $d\ge3$, there is a mathematically precise sense
in which one recovers the original Gleason theorem asymptotically.
Specifically, if a set of QIVPMs representing bounded resources
correspond individually to at least one state's~$\rho'$, those states
will converge to a unique state $\rho$ when the size of the intervals
characterizing the uncertainties tends to zero, and the set of
single-point QIVPMs densely covers the entire real interval $[0,1]$.

@@

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain
events~\cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}. In the
quantum case, the events of interest are specified by \emph{projection
  operators} $P$. These include the empty projector~$\mathbb{0}$, the
identity projector~$\mathbb{1}$, projectors of the form $\proj{\phi}$
where $\ket{\phi}$ is a pure quantum state (an element of a Hilbert
space), sums of \emph{orthogonal} projectors $P_0$ and~$P_1$ with
$P_0P_1=\mathbb{0}$, and products of \emph{commuting} projectors $P_0$
and~$P_1$ with $P_0P_1=P_1P_0$. In a quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event~$P_{i}$ is mapped to a probability~$\mu(P_{i})$ using a
probability measure~$\mu:\events\rightarrow[0,1]$, where $\events$ is
the set of all projectors on a given Hilbert space, subject to the
following constraints: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for each pair
of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
{\mu}\left(P_{0}+P_{1}\right)={\mu}\left(P_{0}\right)+{\mu}\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}
Given a probability for every projector, we can associate an
expectation value with an observable~$\mathbf{O}$ with spectral
decomposition $\mathbf{O}=\sum_{i=1}^{N}\lambda_{i}P_{i}$ as:
\begin{equation}
\expval{\mathbf{O}}_{\mu}=\sum_{i=1}^{N}\lambda_{i}\mu(P_{i})\,.\label{eq:quantum-expectation}
\end{equation}

A conventional quantum probability measure can be easily
constructed if one knows the current state of the quantum system by
using the Born rule.  Specifically, for each pure normalized quantum
state~$\ket{\phi}$, the Born rule induces a probability measure
$\mu_{\phi}^{\mathrm{B}}$ defined as
$\mu_{\phi}^{\mathrm{B}}(P)=\ip{\phi}{P\vert\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$ in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{\mathrm{B}}$ defined as
$\mu_{\rho}^{\mathrm{B}}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\mu_{\phi_{j}}^{\mathrm{B}}\left(P\right)$~\cite{peres1995quantum,544199,Jaeger2007}.

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable
$\mathbf{O}$ with spectral decomposition
$\mathbf{O} = \proj{0} + 2\,\proj{1} + 3\,\proj{2}$. Two fragments of
valid probability measures $\mu_1$ and $\mu_2$ that can be associated
with this space are: $\mu_1(\proj{0}) = 0.5$, $\mu_1(\proj{1}) = 0.5$,
$\mu_1(\proj{2}) = 0$, and $\mu_2(\proj{0}) = 0.5$,
$\mu_2(\proj{1}) = 0$, $\mu_2(\proj{2}) = 0.5$. By the Born rule, the
first probability measure corresponds to the quantum system being in
the pure state $\ket{\phi_1}=(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the
second corresponds to the quantum system being in the pure state
$\ket{\phi_2}=(\ket{0}+\alpha\ket{2})/\sqrt{2}$, where in both cases
$|\alpha|=1$. The expectation values of the observable $\mathbf{O}$,
$\expval{\mathbf{O}}_{\mu_{1,2}}$, are 1.5 in the first case and 2 in
the second.

The above axioms describe a mathematical idealization involving
infinitely precise measurements. In an actual experimental setup with
an ensemble of quantum states that are not exactly identically
prepared, and with small imperfections and inaccuracies in measuring
devices, an experimenter might only be able to determine that the
probability of the event $\proj{0}$ is in the range $[0.49,0.51]$
instead of precisely 0.5. The spread in this range depends on the
amount of resources (time, energy, money, etc.) that are devoted to
the experiment. In the classical setting, this ``fuzziness'' can be
formalized by moving to \emph{interval-valued probability measures}
(IVPMs), which we explore in the next section, along with our proposed
extension to the quantum domain.

@@

We will start by reviewing IVPMs in the classical setting and then
propose our quantum generalization. In the classical setting, there
are several proposals for ``imprecise
probabilities''~\citep{Dempster1967,Shafer1976,GilboaSchmeidler1994,Marinacci1999,Weichselberger2000,JamisonLodwick2004,HuberRonchetti2009,Grabisch2016}.
Although these proposals differ in some details, they all share the
fact that the probability $\mu(E)$ of an event~$E$ is generalized from
a single \emph{real number} to an \emph{interval}~$[\ell,r]$
where~$\ell$ intuitively corresponds to the strength of evidence for
the event~$E$ and~$1-r$ corresponds to the strength of evidence
against the same event. The collection of intervals used in the
definition of a probability measure must satisfy certain properties
that we explain next.

First, for each interval $[\ell,r]$ we have the natural constraint
$0 \leq \ell \leq r \leq 1$ that guarantees that every element of the
interval can be interpreted as a conventional probability. We also
include $\imposs=[0,0]$ and $\necess=[1,1]$ as limiting intervals to
refer to the probability of impossible events (e.g., the null event)
and events that are certain (e.g., the event covering the entire
sample space).  For each interval $[\ell,r]$, we also need the dual
interval $[1-r,1-\ell]$ so that if one interval refers to the
probability of an event~$E$, the dual refers to the probability of the
event's complement $\overline{E}$.  For example, if we discover as a
result of an experiment that $\mu(E) = [0.2,0.3]$ for some event~$E$,
we may conclude that $\mu(\overline{E}) = [0.7,0.8]$ for the
complement event $\overline{E}$. In addition to these simple
conditions, there are some subtle conditions on how intervals are
combined, which we discuss next.

Let $E_1$ and $E_2$ be two disjoint events with probabilities
$\mu(E_1)=[\ell_1,r_1]$ and $\mu(E_2)=[\ell_2,r_2]$. A first attempt
at calculating the probability of the combined event that
\emph{either} $E_1$ or $E_2$ occurs might be
$\mu(E_1\cup E_2) = [\ell_1+\ell_2,r_1+r_2]$. In some cases, this is
indeed a sensible definition. For example, if $\mu(E_1)=[0.1,0.2]$ and
$\mu(E_2)=[0.3,0.4]$ we get $\mu(E_1\cup E_2) = [0.4,0.6]$. But
consider an event $E$ such that $\mu(E)=[0.2,0.3]$ and hence
$\mu(\overline{E})=[0.7,0.8]$. The two events $E$ and $\overline{E}$
are disjoint; the naïve addition of intervals would give
$\mu(E\cup\overline{E})=[0.9,1.1]$ which is not a valid probability
interval. Moreover the event $E\cup\overline{E}$ is the identity event
covering the entire space; its probability interval should be
$\necess$ which is sharper than $[0.9,1.1]$. The problem is that the
two intervals are correlated: there is more information in the
combined event than in each event separately. In our example, even
though the ``true'' probability of $E$ can be anywhere in the range
$[0.2,0.3]$ and the ``true'' probability of~$\overline{E}$ can be in
anywhere in the range $[0.7,0.8]$, the values are not independent. Any
value of $\mu(E) \leq 0.25$ will force $\mu(\overline{E})\geq
0.75$. To account for such subtleties, the axioms of interval-valued
probability do not use a strict equality for the combination of
disjoint events. The correct constraint regarding the probability of
$E_1\cup E_2$ when $E_1$ and $E_2$ are disjoint is
$\mu(E_1\cup E_2) \subseteq [\ell_1+\ell_2,r_1+r_2]$ which, in
combination with the requirement for the identity event would force
$\mu(E \cup \overline{E})$ to always be $\necess$ irrespective of the
particular assignments of $\mu(E)$ or~$\mu(\overline{E})$.

@@

We conclude this section with a generalization of expectation values
of observables in the context of QIVPMs. In conventional quantum
mechanics the expectation value of an observable as defined in
Eq.~\eqref{eq:quantum-expectation} is a unique real number. The
generalization to QIVPMs implies that this expectation value will
itself become bounded by an interval.

\begin{definition}[Expectation Value of Observables over QIVPMs] Let
  $\mathscr{I}$ be a set of intervals; $\Hilb$ a Hilbert space
  with event space~$\events$; and $\mathbf{O}$ an
  observable with spectral decomposition
  $\Sigma_{i=1}^N \lambda_iP_i$. Let $\events'$ be the minimal
  subspace of events containing all the projectors $P_i$ in the
  spectral decomposition of $\mathbf{O}$, then we define:
\begin{equation}
\expval{\mathbf{O}}_{\bar{\mu}}=\left[\min_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}},\max_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}}\right]\,. \label{eq:quantum-interval-expectation-core}
\end{equation}
\end{definition}

\noindent Intuitively the expectation value of an observable relative
to a QIVPM $\bar{\mu}$ lies between two possible outcomes, which
themselves lie between the minimum and maximum bounds of the
probability intervals associated with each state~$\rho$ that is
consistent with $\bar{\mu}$ on every projector in the spectral
decomposition of the observable. If $\bar{\mu}$ is a conventional
(Born) probability measure induced by a state $\rho$, then
$\coreBorn\left(\bar{\mu},\events'\right)$ will contain the singleton
state $\rho$ and both endpoints of the interval will agree, thus
reducing the definition to that of
Eq.~(\ref{eq:quantum-expectation}). We also note that, when restricted
to commuting projectors, the definition above is consistent with the
classical notion of the \emph{Choquet
  integral}~\cite{GilboaSchmeidler1994,Grabisch2016} which is used to
calculate the expectation value of random variables as a weighted
average.

>>>>

We also investigate, though with less insight-generating results, the
second key aspect of imprecise quantum measurement, which is its
statistical nature as reflected in the application of the Born rule
determining the probability of a given measurement
outcome~\citep{Born1983,peres1995quantum,544199}.  Here, Gleason's
theorem establishes that there is no alternative to the Born rule by
demonstrating, using reasonable continuity arguments in Hilbert spaces
of dimension $d\ge3$, the existence of a unique state $\rho$
consistent with the statistical predictions computed from that
rule~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.  Our
question then concerns what happens when infinite precision cannot be
achieved.  We are able to show that, while a QIVPM incorporating the
effects of finite precision might not be consistent with Gleason's
unique state~$\rho$ on all projectors defined on a Hilbert space
$\Hilb$ of dimension $d\ge3$, there is a mathematically precise sense
in which one recovers the original Gleason theorem asymptotically.
Specifically, if a set of QIVPMs representing bounded resources
correspond individually to at least one state's~$\rho'$, those states
will converge to a unique state $\rho$ in the limit as the size of the intervals
characterizing the uncertainties goes to zero, and the set of
single-point QIVPMs densely covers the entire real interval $[0,1]$ or the
appropriate subset corresponding to a given quantum system.

@@

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain
events~\cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}. In the
quantum case, the events of interest are specified by \emph{projection
  operators} $P$. These include the empty projector~$\mathbb{0}$, the
identity projector~$\mathbb{1}$, projectors of the form $\proj{\phi}$
where $\ket{\phi}$ is a pure quantum state (an element of a Hilbert
space), sums of \emph{orthogonal} projectors $P_0$ and~$P_1$ with
$P_0P_1=\mathbb{0}$, and products of \emph{commuting} projectors $P_0$
and~$P_1$ with $P_0P_1=P_1P_0$. In a quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event~$P_{i}$ is mapped to a probability~$\mu(P_{i})$ using a
probability measure~$\mu:\events\rightarrow[0,1]$, where $\events$ is
the set of all projectors on a given Hilbert space, subject to the
following constraints: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for each pair
of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
{\mu}\left(P_{0}+P_{1}\right)={\mu}\left(P_{0}\right)+{\mu}\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}
Given a probability for every projector, we can define the
expectation value of an observable~$\mathbf{O}$ having spectral
decomposition $\mathbf{O}=\sum_{i=1}^{N}\lambda_{i}P_{i}$ as:
\begin{equation}
\expval{\mathbf{O}}_{\mu}=\sum_{i=1}^{N}\lambda_{i}\mu(P_{i})\,.\label{eq:quantum-expectation}
\end{equation}

A conventional quantum probability measure can easily be
constructed using the Born rule if one knows the current state of the quantum
system.  Specifically, for each pure normalized quantum
state~$\ket{\phi}$, the Born rule induces a probability measure
$\mu_{\phi}^{\mathrm{B}}$ defined as
$\mu_{\phi}^{\mathrm{B}}(P)=\ip{\phi}{P\vert\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$, in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{\mathrm{B}}$ defined as
$\mu_{\rho}^{\mathrm{B}}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\mu_{\phi_{j}}^{\mathrm{B}}\left(P\right)$~\cite{peres1995quantum,544199,Jaeger2007}.

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable
$\mathbf{O}$ with spectral decomposition
$\mathbf{O} = \proj{0} + 2\,\proj{1} + 3\,\proj{2}$, i.e., $q_{j} = j$ and
$\ket{\phi_{j}} = \ket{j - 1}$.
Two fragments of valid probability measures $\mu_1$ and $\mu_2$ that can be associated
with this space are: $\mu_1(\proj{0}) = 0.5$, $\mu_1(\proj{1}) = 0.5$,
$\mu_1(\proj{2}) = 0$, and $\mu_2(\proj{0}) = 0.5$,
$\mu_2(\proj{1}) = 0$, $\mu_2(\proj{2}) = 0.5$. By the Born rule, the
first probability measure corresponds to the quantum system being in
the pure state $\ket{\phi_1}=(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the
second corresponds to the quantum system being in the pure state
$\ket{\phi_2}=(\ket{0}+\alpha\ket{2})/\sqrt{2}$, where in both cases
$|\alpha|=1$. The expectation values of the observable $\mathbf{O}$,
$\expval{\mathbf{O}}_{\mu_{1,2}}$, are 1.5 in the first case and 2 in
the second.

The above axioms describe a mathematical idealization involving
infinitely precise measurements. In an actual experimental setup with
an ensemble of quantum states that are not exactly identically
prepared, and with small imperfections and inaccuracies in measuring
devices, an experimenter might only be able to determine that the
probability of the event $\proj{0}$ is concentrated in the range $[0.49,0.51]$
instead of precisely 0.5. The spread in this range depends on the
amount of resources (time, energy, money, etc.) that are devoted to
the experiment. In the classical setting, this ``fuzziness'' can be
formalized by moving to \emph{interval-valued probability measures}
(IVPMs), which we explore in the next section, along with our proposed
extension to the quantum domain.

@@

We will start by reviewing IVPMs in the classical setting and then
propose our quantum generalization. In the classical setting, there
are several proposals for ``imprecise
probabilities''~\citep{Dempster1967,Shafer1976,GilboaSchmeidler1994,Marinacci1999,Weichselberger2000,JamisonLodwick2004,HuberRonchetti2009,Grabisch2016}.
Although these proposals differ in some details, they all share the
fact that the probability $\mu(E)$ of an event~$E$ is generalized from
a single \emph{real number} to an \emph{interval}~$[\ell,r]$
where~$\ell$ intuitively corresponds to the strength of evidence for
the event~$E$ and~$1-r$ corresponds to the strength of evidence
against the same event. More generally, this inverval would be interpreted as
the Guassian width of a probability distribution, but for notational convenience
we will continue in the following to write this as a definite interval.
The collection of intervals used in the
definition of a probability measure must satisfy certain properties
that we explain next.

First, for each interval $[\ell,r]$ we have the natural constraint
$0 \leq \ell \leq r \leq 1$ that guarantees that every element of the
interval can be interpreted as a conventional probability. We also
include $\imposs=[0,0]$ and $\necess=[1,1]$ as limiting intervals to
refer to the probability of impossible events (e.g., the null event)
and events that are certain (e.g., the event covering the entire
sample space).  For each interval $[\ell,r]$, we also need the dual
interval $[1-r,1-\ell]$ so that if one interval refers to the
probability of an event~$E$, the dual refers to the probability of the
event's complement $\overline{E}$.  For example, if we discover as a
result of an experiment that $\mu(E) = [0.2,0.3]$ for some event~$E$,
we may conclude that $\mu(\overline{E}) = [0.7,0.8]$ for the
complementary event $\overline{E}$. In addition to these simple
conditions, there are some subtle conditions on how intervals are
combined, which we discuss next.

Let $E_1$ and $E_2$ be two disjoint events with probabilities
$\mu(E_1)=[\ell_1,r_1]$ and $\mu(E_2)=[\ell_2,r_2]$. A first attempt
at calculating the probability of the combined event that
\emph{either} $E_1$ or $E_2$ occurs might be
$\mu(E_1\cup E_2) = [\ell_1+\ell_2,r_1+r_2]$. In some cases, this is
indeed a sensible definition. For example, if $\mu(E_1)=[0.1,0.2]$ and
$\mu(E_2)=[0.3,0.4]$ we get $\mu(E_1\cup E_2) = [0.4,0.6]$. But
consider an event $E$ such that $\mu(E)=[0.2,0.3]$ and hence
$\mu(\overline{E})=[0.7,0.8]$. The two events $E$ and $\overline{E}$
are disjoint; the naïve addition of intervals would give
$\mu(E\cup\overline{E})=[0.9,1.1]$ which is not a valid probability
interval. Moreover the event $E\cup\overline{E}$ is the identity event
covering the entire space; its probability interval should be
$\necess$ which is sharper than $[0.9,1.1]$. The problem is that the
two intervals are correlated: there is more information in the
combined event than in each event separately. In our example, even
though the ``true'' probability of $E$ can be anywhere in the range
$[0.2,0.3]$ and the ``true'' probability of~$\overline{E}$ can be in
anywhere in the range $[0.7,0.8]$, the values are not independent. Any
value of $\mu(E) \leq 0.25$ will force $\mu(\overline{E})\geq
0.75$. To account for such subtleties, the axioms of interval-valued
probability do not use a strict equality for the combination of
disjoint events. The correct constraint regarding the probability of
$E_1\cup E_2$ when $E_1$ and $E_2$ are disjoint is
$\mu(E_1\cup E_2) \subseteq [\ell_1+\ell_2,r_1+r_2]$ which, in
combination with the requirement for the identity event would force
$\mu(E \cup \overline{E})$ to always be $\necess$ irrespective of the
particular assignments of $\mu(E)$ or~$\mu(\overline{E})$. Parallel arguments
can be made interpreting $[\ell,r]$ as Gaussian width, but will be omitted as no
essential new information seems to be introduced.

@@

We conclude this section with a generalization of expectation values
of observables in the context of QIVPMs. In conventional quantum
mechanics the expectation value of an observable as defined in
Eq.~\eqref{eq:quantum-expectation} is a unique real number. The
generalization to QIVPMs implies that this expectation value will
itself become bounded by or, more generally distributed around, an interval.

\begin{definition}[Expectation Value of Observables over QIVPMs] Let
  $\mathscr{I}$ be a set of intervals; $\Hilb$ a Hilbert space
  with event space~$\events$; and $\mathbf{O}$ an
  observable with spectral decomposition
  $\Sigma_{i=1}^N \lambda_iP_i$. Let $\events'$ be the minimal
  subspace of events containing all the projectors $P_i$ in the
  spectral decomposition of $\mathbf{O}$ and define:
\begin{equation}
\expval{\mathbf{O}}_{\bar{\mu}}=\left[\min_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}},\max_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}}\right]\,. \label{eq:quantum-interval-expectation-core}
\end{equation}
\end{definition}

\noindent Intuitively the expectation value of an observable relative
to a QIVPM $\bar{\mu}$ lies between two possible outcomes, which
themselves lie between the minimum and maximum bounds of the
probability intervals associated with each state~$\rho$ that is
consistent with $\bar{\mu}$ on every projector in the spectral
decomposition of the observable. If $\bar{\mu}$ is a conventional
(Born) probability measure induced by a state $\rho$, then
$\coreBorn\left(\bar{\mu},\events'\right)$ will contain the singleton
state $\rho$ and the interval collapses to a point, thus
reducing the definition to that of
Eq.~(\ref{eq:quantum-expectation}). We also note that, when restricted
to commuting projectors, the definition above is consistent with the
classical notion of the \emph{Choquet
  integral}~\cite{GilboaSchmeidler1994,Grabisch2016} which is used to
calculate the expectation value of random variables as a weighted
average.
